---
title: "Toby's website"
author: "Tobias Schwoerer"
date: "August 16, 2018"
output: html_document
---

```{r setup, include=T}
knitr::opts_chunk$set(echo = F)
```

```{r}
suppressPackageStartupMessages({
  library(leaflet)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(DT)
})
```

# Load salmon data from the knb, then use the local copy and if not present get it from KNB repository 
import the data from knb, but once it is downloaded we can use it from the local computer, this code looks for a local copy of the data and if it is not there, it will download the data. 
```{r}
data_url <- "https://knb.ecoinformatics.org/knb/d1/mn/v2/object/knb.92020.1"
# data_url <- "https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3Af119a05b-bbe7-4aea-93c6-85434dcb1c5e"

esc <- tryCatch(
    read.csv("data_raw/escapement.csv", stringsAsFactors = FALSE),
    error=function(cond) {
        message(paste("Escapement file does not seem to exist, so get it from the KNB."))
        esc <- read.csv(url(data_url, method = "libcurl"), stringsAsFactors = FALSE)
        return(esc)
    }
)

head(esc)
```

```{r}
median_esc <- esc %>% 
  separate(sampleDate, c("Year", "Month", "Day"), sep = "-") %>% 
  # pulls the hyphen in year month and day and turns it into year column, month, and day
  group_by(Species, SASAP.Region, Year, Location) %>% 
  summarize(escapement = sum(DailyCount)) %>% 
  group_by(Species) %>% 
  summarize(median_escapement = median(escapement))

head(median_esc)
```


#Plot the escapement as a bar chart

```{r}
library(cowplot)
ggplot(median_esc, aes(Species, median_escapement)) +
  geom_col() +
  coord_flip()

```

# Now work on building a map
We are starting from daily escapement data again, and finding the unique locations at which escapement was measured. 

## First find unique location points

Make sure to remember to deal with missing vlaues in the location field. 
```{r}
location <- esc %>% 
  distinct(Location, Latitude, Longitude) %>% 
  drop_na()
datatable(location)
```

## Draw the map
Using leaflet
```{r}
leaflet(location) %>% 
  addTiles() %>% 
  addMarkers(~ Longitude, ~ Latitude, popup = ~ Location)

```

Some of the coordinates were wrong entries, need to fix
```{r}
locs <- location %>% mutate(Longitude = abs(Longitude) * -1)

leaflet(locs) %>% 
  addTiles() %>% 
  addMarkers(~ Longitude, ~ Latitude, popup = ~ Location)
```

